Emergent dual scaling of riverine biodiversity

### 2021-03-25 manuscript_pnas.Rmd
Update manuscript_pnas.Rmd

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions to show a robust scaling law of riverine biodiversity along the two orthogonal dimensions of ecosystem size and complexity. In river networks, the recurrent merging of various tributaries forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. The relationships were linear on logarithmic axes, indicating a power-law scaling by ecosystem size and complexity. In support of this theoretical prediction, the dual power-law scaling of riverine fish diversity has consistently emerged across the study regions (Hokkaido Island in Japan and the midwestern United States) despite hosting different fauna with distinct evolutionary histories. The emergence of dual scaling law may be a pervasive property of branching networks with important implications for biodiversity conservation.

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar scaling in $\gamma$ diversity emerge through different ecological pathways.

>>>>

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions to show that riverine biodiversity follows a robust scaling law along the two orthogonal dimensions of ecosystem size and complexity. In river networks, the recurrent merging of various tributaries forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. The relationships were linear on logarithmic axes, indicating a power-law scaling by ecosystem size and complexity. In support of this theoretical prediction, the dual power-law scaling of riverine fish diversity has consistently emerged across the study regions (Hokkaido Island in Japan and the midwestern United States) despite hosting different fauna with distinct evolutionary histories. The emergence of dual scaling law may be a pervasive property of branching networks with important implications for biodiversity conservation.

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar scaling in $\gamma$ diversity emerges through different ecological pathways.
